LME-MPM applied to quasi-brittle fracture.

### 2020-04-21 2020_EFM_MPM_Eigensoftening.tex
Version 21-4-2020

The simulation of fracture propagation in a more accurate and
effective way can be considered as one of the original drivers for
developing novel spatial discretization methods such as meshfree
methods like the material point method (MPM). Presence of cracks are a violation of the
continuity requirement of the finite element approach (MPM).

The MPM does not suffer from the above difficulties. Discontinuities
can be described in two ways. One is to abandon the single-valued
velocity field property near the crack by using two or more background
meshes, and the other is to use failed material points to
approximately describe the crack.

In the first approach,

@@

The popularity of MPM has increase notoriously during the recent years
due to its ability to deal with large strain problems without mesh
distorsion issues inherent to mesh based methods like FEM, see
Zdzislaw \cite{Wieckowski2004}. However, in the simulations made with
the original MPM, there are numerical noises when particles crossing
the cell boundaries.

@@

For the spatial discretization, two sets of points are introduced in
the MPM. First, the nodes, this points are considered fixed in the space and are in charge of
computing all the kinematic fields such forces $f_I$, accelerations $a_I$ and
velocities $v_I$. And second the material points or particles. They are in
charge of the discretization of the continuum, and store the local state ($\sigma_p, \varepsilon_p$).

@@

\subsection{Temporal discretization}
\label{sec:2.3}

This research is devoted to capture the challenging process of
fracture during high velocity impacts. To capture the presence of
elastic shock waves triggered by the fracture process. Here we adopt
a explicit predictor-corrector time integration
scheme. It is based in the Newmark a-form 
$\gamma = 0.5$ and $\beta = 0$ which is the central difference
explicit. In a first stage, the nodal velocity predictor is computed
following \eqref{eq:Predictor-velocity}, 
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{\tilde{v}}_I^{k+1} = \frac{ N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{m_I}
\end{equation}
This way of computing the nodal predictor is both numerically stable
and minimize the computational effort. Once nodal velocity are
obtained, the essential boundary conditions are imposed. And in the
following, the ``classic'' MPM algorithm continues to reach to the
equilibrium equation \eqref{eq:particle_balance_forces3}. Here we
continue with the \textit{corrector} stage, due to the fact that we
already have nodal velocity, this step is computed in the same way as
in FEM,
\begin{equation}
  \label{eq:Corrector-velocity}
  \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\ \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}
\end{equation}
Finally updated particle kinetics are computed using nodal values as,
\begin{align}
  \label{eq:Update-lagrangian-pce}
        &\vec{a}_p^{k+1} = \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k} 
\end{align}
The complete pseudo-algorithm it is summarized in \ref{sec:expl-pred-corr}.

\subsection{Fracture modelling approach}
\label{sec:2.4}

@@

As can be noticed, in the eigenerosion algorithm an energetic
criterion is adopted. Due to that fact, unrealistic stress
concentration (higher than tensile strength) in quasi-brittle materials, see
\cite{Navas_2017_ES}. To overcome this limitation, the aforementioned
authors proposed the concept of eigensoftening to take in to account
the gradual failure in quasi-brittle materials. The concept is
inspired in the cohesive fracture widely employed in the context of
FEM \cite{Ortiz_1999}. This gradual failure criterion is plotted in
figure \label{fig:Damage-ft-wc}, where a linear decreasing cohesive
law is presented to illustrate the concept here described. In the
picture, the shaded area represents the static fracture energy per
unit of area, $G_F$. As we can see, a cohesive crack appears when the
maximum tensile strength, $f_t$ is reached. Once the opening
displacement $w$ takes the value of the critical crack displacement
$w_c$, a stress-free crack is attained. For intermediate values 
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/Damage}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical
    opening displacement.}
  \label{fig:Damage-ft-wc}
\end{figure}
%%%%%%%%%%%

>>>>

The simulation of fracture propagation in a more accurate and
effective way can be considered as one of the original drivers for
developing novel spatial discretization methods such as meshfree
methods like the material point method (MPM). Presence of cracks are a
violation of the continuity requirement of the finite element method
(FEM).

By contrast, MPM does not suffer from the above difficulty since the
continuity requirement is less restrictive. Regarding fracture, it can
be described in two ways. One is to remove the restriction of the single-valued velocity field close to the crack by using two or
more sets of nodes \cite{Nairn_2006}. This method is based in to
assign different labels to distinguish if the material points and
nodes are in the same side of the crack or not. Under this approach
the crack surface is described with line segments in 2D and triangle
patches in 3D cases. This method was named as ``CRAcks with Material
Points (CRAMP)'' \cite{Nairn_2003}. In this method, the criteria for
crack propagation is based on such parameters as energy release rate
analyzed by Tan \& Nairn (2002)\cite{Nairn_2002}, and the stress
intensity factor or the J-integral discussed by Guo \& Nairn
(2004)\cite{Nairn_2004}.

and the other is to use failed material points to
describe the crack evolution.

The erosion of the material point means that each material point can
be either intact or be completely failed or eroded and has no load
bearing capacity, Even thought the method has been successfully
applied to dynamic fragmentation of metals, quantitative validations,
such as stress or strain levels near or at the crack set compared to
experimental measures are still lacking. Having in mind that the
eigenerosion approach was intended for perfectly brittle fracture, an
eigensoftening concept is developed for quasi-brittle materials

@@

The aim of this section is to provide an overview of the special
techniques employed to face the fracture problem under the MPM
framework. In consequence it is structured as follows: first in
\ref{sec:2.1} we summarize the explicit-predictor algorithm for the
MPM, next the local \textit{max-ent} approximants are introduced in
\ref{sec:2.2} as an accurate alternative technique to interpolate data
between particles and nodes, and finally the fracture algorithms based
in the eigendeformations are presented in \ref{sec:2.3}.

@@

The MPM~\cite{Sulsky1994} is a meshfree Lagrangian-Eulerian method
where particles carries on all the physical information and a set of
background nodes is employed to compute the equilibrium
equation. Since the MPM possesses the advantages of both Lagrangian
and Eulerian descriptions, no element distortion exists in the MPM,
therefore it is an appropriate and efficient method in solving
problems involving extremely large deformation and moving
discontinuities such fracture evolution. For the spatial
discretization, two sets of points are introduced in 
the MPM. First, the nodes. This points are considered fixed in the
space and are in charge of computing all the kinematic fields such
forces $f_I$, accelerations $a_I$ and velocities $v_I$. And second the
material points or particles. They are in charge of the discretization
of the continuum, and store the local state ($\sigma_p,
\varepsilon_p$). Without loosing generality, the MPM algorithm can be
described with three main steps: (i) a
variational recovery process, where particle data is projected to the
grid nodes, (ii) an Eulerian step, where balance of momentum equation
is expressed as a nodal equilibrium equation thorough a FEM-like
procedure, and finally (iii) a Lagrangian advection of the
particles. In the present research we will adopt a explicit
predictor-corrector time integration scheme. The purpose of this
choice is motivated due its proved robustness and stability in
numerical calculations. In the first stage (i), the nodal velocity
predictor is computed following \eqref{eq:Predictor-velocity}, 
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{\tilde{v}}_I^{k+1} = \frac{ N_{Ip}^{k} m_p (\vec{v}_p^k + (1 - \gamma)\ \Delta t\ \vec{a}_p^k)}{m_I}
\end{equation}
This way of computing the nodal predictor is both numerically stable
and minimize the computational effort. Once nodal velocity are
obtained, the essential boundary conditions are imposed. And in the
following, a Eulerian phase (ii) is computed in the set of nodes in a
FEM-like way, where nodal forces $\vec{f}_{I}^{k+1}$ are computed thorough the
equilibrium equation. Next the nodal velocities are corrected in a
\textit{corrector} stage,
\begin{equation}
  \label{eq:Corrector-velocity}
  \vec{v}_{I}^{k+1} = \vec{v}_{I}^{pred} + \gamma\ \Delta t\ \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}
\end{equation}
Finally updated the particles are advected in the Lagrangian stage (iii) using nodal values as,
\begin{align}
  \label{eq:Update-lagrangian-pce}
        &\vec{a}_p^{k+1} = \frac{N_{Ip}^k\vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
         N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k} 
\end{align}
The complete pseudo-algorithm it is summarized in \ref{sec:expl-pred-corr}.

@@

\subsection{Fracture modelling approach}
\label{sec:2.3}

@@

As can be noticed, in the eigenerosion algorithm an energetic
criterion is adopted. Due to that fact, unrealistic stress
concentration (higher than tensile strength) in quasi-brittle materials, see
\cite{Navas_2017_ES}. To overcome this limitation, the aforementioned
authors proposed the concept of eigensoftening to take in to account
the gradual failure in quasi-brittle materials. The concept is
inspired in the cohesive fracture widely employed in the context of
FEM \cite{Ortiz_1999}. This gradual failure criterion is plotted in
figure \label{fig:Damage-ft-wc}, where a linear decreasing cohesive
law is presented to illustrate the concept here described. In the
picture, the shaded area represents the static fracture energy per
unit of area, $G_F$. As we can see, a cohesive crack appears when the
maximum tensile strength, $f_t$ is reached. Once the opening
displacement $w$ takes the value of the critical crack displacement
$w_c$, a stress-free crack is attained. For intermediate values,
$w_n$, a damage value between zero and one represents the extension to
which the material has failed.
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figures/Damage}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical
    opening displacement.}
  \label{fig:Damage-ft-wc}
\end{figure}
%%%%%%%%%%%
For the eigensoftening algorithm,

a strength criterion for crack initialization was adopted
